---
title: "Getting Started with mmfpx"
author: "Josh K. Lee"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mmfpx}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette walks through a minimal workflow for the `mmfpx` package using the
included `simulated_longitudinal_data` dataset.

`mmfp()` fits fractional polynomial mixed-effects models so you can uncover
non-linear age trajectories while still accounting for subject-level
correlation. It handles the common problem of longitudinal biomarkers whose
curvature is missed by standard linear random-effects models.
By default it keeps the raw age scale (after verifying all ages are strictly
positive), but you can optionally request ratio scaling via
`centering = "ratio"` or variance-stabilising scaling with
`centering = "sd"`, which divides age by its sample standard deviation without
recentering.
By default `mmfp()` evaluates both one-term (FP1) and two-term (FP2) models;
set `fp_models = "fp1"` when you only want one-term candidates or
`fp_models = "fp2"` when you only want two-term candidates.

`normbytcv()` rescales regions of interest against a reference volume to remove
global size effects. It addresses the need to compare baseline scans from
subjects with different head sizes or tissue totals when longitudinal
measurements are not the focus.

## Data overview

```{r}
library(mmfpx)

data("simulated_longitudinal_data")

str(simulated_longitudinal_data)
```

For the examples below we use the full dataset, which remains manageable for
the models demonstrated here.

## Fractional polynomial mixed-effects models

We can evaluate candidate fractional polynomial mixed-effects models for
`volumeA` as the outcome, using `age` as the predictor and `subj_id` for the
random-effects grouping structure.

```{r}
fp_results <- mmfp(
  data = simulated_longitudinal_data,
  outcome_vars = "volumeA",
  age_var = "age",
  id_var = "subj_id",
  visit_var = "visit",
  powers = c("log", 1),
  random_intercept_only = TRUE
)

fp_results$results$volumeA$summary

fp_results$results$volumeA$selection_summary
```

To fit only one-term fractional polynomial models, set `fp_models = "fp1"`:

```{r}
fp_results_fp1_only <- mmfp(
  data = simulated_longitudinal_data,
  outcome_vars = "volumeA",
  age_var = "age",
  id_var = "subj_id",
  visit_var = "visit",
  powers = c("log", 1),
  fp_models = "fp1",
  random_intercept_only = TRUE
)

fp_results_fp1_only$results$volumeA$summary
```

To fit only two-term fractional polynomial models, set `fp_models = "fp2"`:

```{r}
fp_results_fp2_only <- mmfp(
  data = simulated_longitudinal_data,
  outcome_vars = "volumeA",
  age_var = "age",
  id_var = "subj_id",
  visit_var = "visit",
  powers = c("log", 1),
  fp_models = "fp2",
  random_intercept_only = TRUE
)

fp_results_fp2_only$results$volumeA$summary
```

The summary table includes log-likelihood, AIC, BIC, `delta_AIC`, Akaike
weights, evidence ratios, convergence, and singular-fit indicators for each
fitted model. Weights are relative support within the fitted candidate set, not
probabilities that a model is correct. The selection summary gives a compact
view of whether the winning model is clearly separated from nearby candidates.
If you encounter convergence challenges or want more interpretable random
effect variances, set `centering = "sd"` to divide age by its standard
deviation before estimating the candidate models.

### Interacting FP terms with a covariate

You can include interactions between the fractional polynomial (FP) terms and
another covariate using a shorthand in `static_formula`. For example, to allow
different FP trajectories by `diagnosis`, supply `"*diagnosis"` as a character
string:

```{r}
fp_results_interact <- mmfp(
  data = simulated_longitudinal_data,
  outcome_vars = "volumeA",
  age_var = "age",
  id_var = "subj_id",
  visit_var = "visit",
  static_formula = "*diagnosis",  # expands to .fp1*diagnosis (+ .fp2*diagnosis)
  powers = c("log", 1),
  random_intercept_only = TRUE
)

fp_results_interact$results$volumeA$summary
```

In FP1 models, the shorthand adds `.fp1*diagnosis`. In FP2 models it adds both
`.fp1*diagnosis` and `.fp2*diagnosis`. You can mix shorthand with additional
additive terms, e.g. `static_formula = "*diagnosis + visit"` to also include a
main effect of `visit`.

## Normalising volumes with `normbytcv`

We can also normalise a region of interest by another volumetric measure using
the `normbytcv()` function. The example below keeps only the first visit per
participant and adjusts `volumeA` for its non-linear association with `volumeB`.

```{r}
baseline_data <- simulated_longitudinal_data[
  simulated_longitudinal_data$visit == 1,
  ,
  drop = FALSE
]

norm_results <- normbytcv(
  data = baseline_data,
  TBV_var = "volumeB",
  ROI_vars = "volumeA",
  subject_id_vars = "subj_id",
  powers_set = c("log", 0.5, 1),
  proportion = 0.75,
  num_cores = 1,
  plot_models = FALSE,
  save_results = FALSE
)

purrr::map_dfr(
  norm_results$results$volumeA$filtered_models,
  ~ tibble::tibble(
    model_type = .x$model_type,
    p1 = .x$powers$p1,
    p2 = .x$powers$p2,
    AIC = .x$model_aic
  )
)

head(norm_results$adjusted_volumes)
```

The resulting list contains adjusted ROI values and information about the
fractional polynomial models that were retained after applying the bagging
criterion.

## Next steps

You can expand this workflow by adding other outcomes, experimenting with the
available power sets, and comparing random-slope specifications or correlation
structures by adjusting the arguments supplied to `mmfp()`.
